---
title: "Configuration"
description: >
  Configuration for research codes
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Configuration

There are two ways we configure our code, by command-line arguments and with configuration files that have parameters.

Sometimes organizations create methods for storing script parameters in a central database that's specifically oriented for this. We won't cover that here, but you can imagine that having a central store makes it easier to find what was used to run code in the past and record runs now.

## Finding files and databases

We run our code on laptops and on the cluster. Someone else will probably run this code next year. One of the problems with running code in different places, at different times, is that input files are changed and input databases have changed names. We can mitigate that problem by using a single layer of indirection.

### Finding files in the project directory

When you work in a package, different pieces of code can be run from different subdirectories, such as vignettes and scripts. To make this easier, you can reference files relative to the path of the package root. This command gives you the root of a package.

```{r, eval = FALSE}
fs::path_package("safetyguide")
```

For more complicated situations, such as finding paths relative to the root of a git repository, the [rprojroot](https://github.com/r-lib/rprojroot) package has you covered.

```{r, eval = FALSE}
rprojroot::find_root(rprojroot::is_rstudio_project)
rprojroot::find_root(rprojroot::is_git_root)
```


## Parameter Files

When there are more than a few parameters, we can use a parameter file. The [configr package](https://cran.r-project.org/web/packages/config/vignettes/introduction.html) can read parameter files written in [YAML](https://yaml.org) or 
[TOML](https://toml.io). TOML has a more regular syntax, so let's use that below.

Notice that the double-brackets are replaced with previously-defined values.

```{asis, eval = FALSE}
[versions]
pfpr = "201029"
am = "201029"
outvars = "201122_1000"

[roles]
pfpr = "/globalrc/inputs/PfPR_medians/{{pfpr}}"
am = "/globalrc/inputs/AM_medians/{{am}}"
outvars = "/globalrc/outputs/basicr/{{outvars}}"

[parameters]
# These are scientific parameters.
kam = 0.6  # modifies AM to get rho
b = 0.55  # biting
random_seed = 24243299
confidence_percent = 95
pfpr_min = 0.02
pfpr_max = 0.98

[options]
# These affect computation but not output scientific numbers.
blocksize = 16
pngres = 150
```

This parameter file uses the word "role" to refer to the role an input or output file plays for the script. It distinguishes options, which shouldn't affect output scientific values, from parameters, which definitely do affect output scientific values.

The code can read te parameters section file with one command.
```{r readparams, eval = FALSE}
parameters <- configr::read.config(config)[["parameters"]]
```
The result is a named list of parameter values.

If we want to record, later, what paramters were used, we could copy this TOML file into an output data directory.

## Command-line arguments

There are several packages that help parse command-line arguments. A dependable choice is [argparse](https://github.com/trevorld/r-argparse). It looks like this:
```{asis}
> library("argparse")
> parser <- ArgumentParser(description='Process some integers')
> parser$add_argument('integers', metavar='N', type="integer", nargs='+',
+                    help='an integer for the accumulator')
> parser$add_argument('--sum', dest='accumulate', action='store_const',
+                    const='sum', default='max',
+                    help='sum the integers (default: find the max)')
> parser$print_help()
usage: PROGRAM [-h] [--sum] N [N ...]

Process some integers

positional arguments:
  N           an integer for the accumulator

optional arguments:
  -h, --help  show this help message and exit
  --sum       sum the integers (default: find the max)
```


I'll show [docopt](https://github.com/docopt/docopt.R) below, but the same principles hold for both:

- If you parse parameters in a separate function, you can test that function more easily.
- There are input parameters and these are translated into algorithm parameters. Doing this translation can speed up computation.

Here's an example of using docopt, which looks at a formatted usage string in order to guess what your parameters must be.
```{r, eval = FALSE}
arg_parser <- function(args = NULL) {
    doc <- "pr to Rc
Usage:
  rc_kappa.R [options]
  rc_kappa.R (-h | --help)
Options:
  -h --help              Show help.
  --config=<config>      A configuration file.
  --country=<alpha3>     The three-letter country code for a country's outline.
  --outvars=<outversion> Version of output to write.
  --overwrite            Whether to overwrite outputs.
  --years=<year_range>   A range of years to do, as an R range, 2000:2010.
  --cores=<core_cnt>     Tell it how many cores to use in parallel.
  --draws=<draw_cnt>     How many draws to use.
  --task=<task_id>       If this is a task, which task.
  --tasks=<task_cnt>     Total number of tasks. You should set this for workers.
"
    if (is.null(args)) {
        args <- commandArgs(TRUE)
    }
    parsed_args <- docopt::docopt(doc, version = "rc_kappa 1.0", args = args)
    if (is.null(parsed_args$config)) {
        parsed_args$config <- "rc_kappa.toml"
    }
    parsed_args
}
```
This function accepts an `args` argument so that a unit test can pass in candidate versions of `commandArgs()` output and check that they are parsed as expected.

It may seem silly to think about testing command-line parameters. I test them because I've had to change parameter values at 3 AM before submitting cluster jobs that would take a week to run. I really want to know I didn't make a typo, before the job launches, so I add a test and run the test suite before submitting the job.

It is often the case that there is a difference between how you want to specify values to a program and what the core algorithm needs. For instance, you may specify a quantile in the input, and then the algorithm works with that quantile of a particular distribution. There is a translation step, and doing it once can save a little processing time. More importantly, it can make later calculation code clearer. This means there can be a parameter translation step early in the code, if you think that helps a particular code base.
